### 2020-06-23 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/mqsi20.tex
Updated code package sent off last night. New tex revisions from Dr. Watson received, implementing.

% Algorithm XXX: MQSI---Monotone quintic spline interpolation,
% Thomas C.H. Lux, Layne T. Watson, Tyler H. Chang,
% William Thacker, Kirk W. Cameron, and Yili Hong
% ACM Trans. Math. Software, ?/?/2020.
% 
%
\magnification=1000 \hsize=30pc \vsize=47pc
\parindent=10pt \parskip=0pt plus 3pt
\baselineskip=13pt plus 2pt minus 1pt \lineskiplimit=2pt
\lineskip=2pt plus 2pt                                  %\raggedright
\tolerance=800
%\abovedisplayskip=6pt plus 6pt minus 3pt
%\abovedisplayshortskip=0pt plus 3pt
%\belowdisplayskip=6pt plus 6pt minus 3pt
%\belowdisplayshortskip=6pt plus 2pt minus 3pt

% eplain package, mainly for its bibliography management.
\input eplain

@@

% Commands for inserting an 'algorithm' section.
\def\indented#1{{\leftskip=1cm}{#1}{\leftskip=0cm}}
\def\comment#1{\vskip 10mm {\it #1} \vskip 10mm}%
\def\thickrule{\hrule\hrule\hrule}

@@

\input epsf %Postscript figures.
\def\lcaption#1#2{{\noindent\rmVIII Fig.~#1. #2\par}}
\def\ltabcaption#1#2{{\noindent\rmVIII Table #1. #2\par}}

@@

\def\leftrh{Lux et al.}
\def\rightrh{Algorithm XXX: MQSI}

{\baselineskip=24pt \leftline{\hlvXVIII Algorithm XXX: MQSI: Monotone Quintic}
\leftline{\hlvXVIII Spline Interpolation}}
\bigskip\medskip
\leftline{\hlv THOMAS C.H. LUX, LAYNE T. WATSON, TYLER H. CHANG,}
\smallskip
\leftline{\hlv WILLIAM THACKER, KIRK W. CAMERON, AND YILI HONG}
\smallskip
\leftline{\hlv Virginia Polytechnic Institute and State University}
\bigskip

\hrule\bigskip\smallskip \footnote{}{\hskip
-\parindent{\parindent=0pt\rmVIII Authors' addresses:
T. C. H. Lux, T. H. Chang, K. W. Cameron,
Department of Computer Science, L. T. Watson,
Departments of Computer Science, Mathematics, and Aerospace and Ocean
Engineering, Y. Hong, Department of Statistics, Virginia
Polytechnic Institute \& State University, Blacksburg, VA 24061;
e-mail: {\ttVIII tchlux@vt.edu}. \xpar
% Permission to make digital/hard copy of
%part or all of this work for personal or classroom use is granted
%without fee provided that the copies are not made or distributed for
%profit or commercial advantage, the copyright notice, the title of
%the publication, and its date appear, and notice is given that
%copying is by permission of the ACM, Inc. To copy otherwise, to
%republish, to post on servers, or to redistribute to lists, requires
%specific permission and/or fee. \xpar\copyright\ 2018 by the
%Association for Computing Machinery, Inc. \xpar 
}}

% Abstract and Introduction
{\rmVIII\parindent=0pt MQSI contains a serial code written in Fortran
  1995 for constructing monotone quintic spline interpolants to
  data. Using sharp theoretical monotonicity constraints, first and
  second derivative estimates at data provided by a quadratic facet
  model are refined to produce a $C^2$ monotone interpolant. This
  paper includes algorithm and implementation details, complexity and
  sensitivity analyses, usage information, and a brief performance
  study.

@@

\bigskip\hrule\bigskip\medskip
\heading{1. INTRODUCTION}

Many domains of science rely on smooth approximations to real-valued
functions over a closed interval. Piecewise polynomial functions
(splines) provide the smooth approximations for animation in graphics
\cite{herman2015techniques,quint2003scalable}, aesthetic structural
support in architecture \cite{brennan2019measure}, efficient
aerodynamic surfaces in automotive and aerospace engineering
\cite{brennan2019measure}, prolonged effective operation of electric
motors \cite{berglund2009planning}, and accurate nonparametric
approximations in statistics \cite{knott2012interpolating}. While
polynomial interpolants and regressors apply broadly, splines are
often a good choice because they can approximate globally complex
functions while minimizing the local complexity of an approximation.

It is often the case that the true underlying function or phenomenon
being modeled has known properties e.g., convexity, positivity,
various levels of continuity, or monotonicity. Given a reasonable
amount of data, it quickly becomes difficult to achieve desirable
properties in a single polynomial function. In general, the
maintenance of function properties through interpolation/regression is
referred to as {\it shape preserving}
\cite{fritsch1980monotone,gregory1985shape}. The specific shapes this
work will achieve in approximations are monotonicity and $C^2$
continuity. These properties are chiefly important to the
approximation of cumulative distribution functions and subsequently
the effective generation of random numbers from a specified
distribution.

In statistics especially, the construction of a monotone interpolating
spline that is $C^2$ continuous is meaningfully useful
\cite{ramsay1988monotone}. A function with these properties could
approximate a cumulative distribution function to a high level of
accuracy with relatively few intervals. A twice continuously
differentiable approximation to a cumulative distribution function
(CDF) would also produce a corresponding probability density function
(PDF) that is continuously differentiable, which is a property many
standard parametric distributions maintain.

The currently available software for monotone piecewise polynomial
interpolation includes quadratic \cite{he1998monotone}, cubic
\cite{fritsch1980monotone}, and (with limited application) quartic
\cite{wang2004rational,piah2011improved,yao2018unconditionally}
cases. In addition, a statistical method for bootstrapping the
construction of an arbitrarily smooth monotone fit exists
\cite{leitenstorfer2006generalized}, but the method does not take
advantage of known analytic properties related to quintic
polynomials. Theory has been provided for the quintic case
\cite{ulrich1994positivity,hess1994positive}, however this theory has
not yet been used to construct a monotone quintic spline interpolation
routine. Recent work suggests that the lack of quintic software may be
due to a general unawareness of the theory
\cite{xie2018semiparametric}.

\topinsert
\centerline{\epsfxsize=4.5truein \epsffile{vt_logo.eps}}
{\narrower\noindent\rmVIII Fig.\ 1.
Example polynomials that interpolate function values at the ends of
the interval $[0,1]$. The first only interpolates the function values
$g_2(0) = 1$ and $g_2(1) = 0$, making it the order two polynomial
$g_2(x) = 1 - x$. For the second plot $g_4(x) = 8x^3 - 12x^2 + 4x +
1/2$, which is order four and interpolates the values $g_4(0) = 1/2$,
$g_4'(0) = 4$, $g_4(1) = 1/2$, $g_4'(1) = 4$. Finally the third plot
shows the order six polynomial $g_6(x) = - 64x^5 + 160x^4 - 140x^3 +
50x^2 - 7x + 1$ interpolating the function values $g_6(0) = 1$,
$g_6'(0) = -7$, $g_6''(0) = 100$, $g_6(1) = 0$, $g_6'(1) = -7$,
$g_6''(1) = -100$. Notice that interpolating the same fixed number of
function values at each endpoint will always result in an even order
interpolating polynomial.
\par}
\endinsert

The importance of piecewise quintic interpolation over lower order
approximations can be simply demonstrated. In general, the order of a
polynomial determines the number of function values it can
interpolate, and the growth rate of error away from the interpolated
function values. As demonstrated in Figure
\ref{fig:spline_demonstration}, it can be seen that matching a value
at either end of the interval requires an order two (linear)
approximation and each additional derivative at the ends of the
interval raises the necessary polynomial order by two. Hence, only
even order (odd degree) interpolating splines are broadly
applicable. The body of this work is composed of a novel algorithm for
enforcing monotonicity on quintic polynomial pieces, then extending
that solution to work on quintic splines.

The major contribution of this work is an algorithm for constructing
monotone quintic interpolating splines that utilizes existing quintic
monotonicity theory. The remainder of this paper is structured as
follows: Section \ref{sec:monotone-cubic} summarizes the existing
monotone cubic spline interpolation methodology, Section
\ref{sec:monotone-quintic} presents an algorithm for constructing
monotone quintic spline interpolants, Section \ref{sec:results} offers
experiments and results with cubic and quintic methods, Sections
\ref{sec:discussion} and \ref{sec:conclusion} discuss results and
conclude.

\heading{1.1 \enspace Computing a Monotone Cubic Interpolant}

The current state-of-the-art monotone interpolating spline with a
mathematical software implementation is piecewise cubic, continuously
differentiable, and was first proposed in \cite{fritsch1980monotone}
then expanded upon in \cite{carlson1985monotone}. Let $\pi: x_0 = k_1
< k_2 < \cdots < k_n = x_1$ be a partition of the interval
$[x_0,x_1]$. Let $f: {\bbb R} \rightarrow {\bbb R}$ be $C^2$, and
$\bigl\{f(k_i)\bigr\}_{i=1}^n$ and $\bigl\{\Delta_i\bigr\}_{i=1}^n$ be
given sets of function and derivative values at the partition points
for a monotone function $f$. Either $f(k_i) \leq f(k_{i+1})$, $i=1$,
$\ldots$, $n-1$, and $\Delta_i\ge0$, $i=1$, $\ldots$, $n$, or $f(k_i)
\geq f(k_{i+1})$, $i=1$, $\ldots$, $n-1$, and $\Delta_i\le0$, $i=1$,
$\ldots$, $n$. Let $\hat f$ be a piecewise cubic polynomial defined in
each subinterval $I_i = [k_i, k_{i+1}]$ by
$$ h_i = k_{i+1} - k_{i}, \quad u(t) = 3t^2 - 2t^3, \quad p(t) = t^3 - t^2,$$
$$\hat f(x) = f(k_i) u\big((k_{i+1} - x) / h_i\big) + f(k_{i+1}) u\big((x - k_i) / h_i\big) $$
$$ - h_i\Delta_i p\big((k_{i+1}-x)/h_i\big) + h_i\Delta_{i+1} p\big((x-k_i)/h_i\big).$$ 
Notice that a trivially monotone spline results when $\Delta_i = 0$,
for $i = 1$, $\ldots$, $n$. However, such a spline has too many
{\it wiggles} for most applications. \cite{carlson1985monotone}
show that simple conditions on the derivative values can guarantee
monotonicity, and that these conditions can be enforced in a way that
ensures modifications on one interval will not break the monotonicity
of cubic polynomials over any neighboring intervals. If $f(k_i) =
f(k_{i+1})$, take $\Delta_i = \Delta_{i+1} =0$ and $\alpha =\beta =1$,
otherwise let $\alpha = \big(\Delta_i (k_{i+1}-k_i)\big) /
\big(f(k_{i+1}) - f(k_i)\big)$ and $\beta =
\big(\Delta_{i+1}(k_{i+1}-k_i)\big) / \big(f(k_{i+1}) -
f(k_i)\big)$. Monotonicity of a cubic polynomial over a subinterval
can be maintained by ensuring that $\alpha$ and $\beta$ reside in any
of the regions depicted in Figure \ref{fig:projection}.

\topinsert
\centerline{\epsfxsize=4.5truein \epsffile{vt_logo.eps}}
{\narrower\noindent\rmVIII Fig.\ 2.
These are the feasible regions of monotonicity for cubic splines and
the projections that make a cubic polynomial piece monotone. The
regions themselves are numbered 1--4 corresponding to their original
description in Fritsch and Carlson (1980). One point is projected onto
each region as a demonstration.
\par}
\endinsert

The actual region of monotonicity for a cubic polynomial is larger,
but projection of $(\alpha, \beta)$ into one of these regions ensures
that monotonicity will be achieved and not violated for neighboring
regions. The user must decide which region is most appropriate for the
projections based on the application, Fritsch and Carlson recommend
using Region 2.

While the cubic monotonicity case affords such a concise solution, the
region of monotonicity is not so simple in the quintic case. In the
next section, an algorithm for performing a projection similar to
those for cubic polynomials is proposed.

\heading{2. MONOTONE QUINTIC INTERPOLATION}

The following section is composed of three algorithms that together
are used to construct a monotone quintic spline interpolant. Without
loss of generality, the algorithms will only consider the monotone
increasing (nondecreasing) case. The monotone decreasing case is
handled similarly. Algorithm \ref{alg:check-monotone} checks
monotonicity, Algorithm \ref{alg:make-monotone} enforces monotonicity
on an order six polynomial piece of the quintic spline, and Algorithm
\ref{alg:monotone-spline} uses the previous two algorithms to enforce
monotonicity for the entire quintic spline.

In pseudocode, a function {\tt binary\_search($g,$ $a,$ $b$)} is
used, where $a,b\in S\subset {\bbb R}^p$ for convex $S$, $g: S
\rightarrow \{${\tt FALSE}, {\tt TRUE}$\}$ is a right continuous
Boolean function, $g(b) = \hbox{{\tt TRUE}}$, and for $\mu \in
[0,1]$, $g\big((1-\mu)a + \mu b\big) =\hbox{{\tt TRUE}}
\Longrightarrow g\big((1-\nu)a + \nu b\big) =\hbox{{\tt TRUE}}$ for
$\mu \le \nu\le1$. The search returns $\big((1-c)a + c b\big)$ for the
smallest $c \in [0,1]$ such that $g\big((1-c)a + c b\big) =
\hbox{{\tt TRUE}}$.

These algorithms make use of first and second derivatives of the approximated function where $D$ denotes the differentiation operator. In the case that the first and second derivative information is not provided along with function values, the derivatives are estimated with finite differences of the function values. The final quintic spline is represented as a piecewise polynomial using the Newton form for each polynomial piece, and ultimately converted to a $B$-spline representation for evaluation.

\def\newline{\hfil\break}

\parindent=0pt
\vskip 5mm
\thickrule
{\bf Algorithm 1:} {\tt is\_monotone}$\big(x_0, x_1, f \big)$ \newline
where $x_0$, $x_1 \in {\bbb R}$, $x_0 < x_1$, and $f$ is an order six
polynomial defined by $f(x_0)$, $Df(x_0)$, $D^2f(x_0)$, $f(x_1)$,
$Df(x_1)$, $D^2f(x_1)$. Returns {\tt TRUE} if $f$ is monotone
increasing on $[x_0,x_1]$.
\hrule
\vskip 3mm
\numberedlist
{\tt
\li if $\big(f(x_0) = f(x_1)\big)$ and not $\big( 0 = Df(x_0) = Df(x_1) = D^2f(x_0) = D^2f(x_1) \big)$
\li \indented{return FALSE} \newline
end if \newline
\li if $\big(Df(x_0) < 0$ or $Df(x_1) < 0\big)$ \newline
    \indented{return FALSE} \newline
    end if \newline
    \comment{The necessity of these first two steps follows directly from the fact that $f$ is $C^2$. The next case is in accordance with a simplified condition for quintic monotonicity that reduces to one of cubic positivity studied in \cite{schmidt1988positivity}, where $\alpha$, $\beta$, $\gamma$, and $\delta$ are defined in terms of values and derivatives of $f$ at $x_0$ and $x_1$. Step \ref{alg1-alpha} checks for the necessary condition that $\alpha \ge 0$, Step \ref{alg1-beta} checks $\beta \geq \alpha$, and Step \ref{alg1-gamma} checks $\gamma \geq \delta$, all from \cite{schmidt1988positivity}. If all necessary conditions are met, then the order six piece is monotone and Step \ref{alg1-simplified-true} concludes this check.}
  if $\big(Df(x_0) = 0$ or $Df(x_1) = 0\big)$ \newline
  \indented{$w := x_0 - x_1$} \newline
  \indented{$v := f(x_0) - f(x_1)$} \newline
  \indented{if $\big(D^2f(x_1) > {-4}Df(x_1) / w \big)$ return FALSE} \newline
  \indented{if $\big(D^2f(x_1) < (3w D^2f(x_0) - 24 Df(x_0) - 32
    Df(x_1) + 60v/w) / (5w) \big)$ return FALSE} \newline
  \indented{if $\big(D^2f(x_0) < 3 Df(x_0) / w \big)$ return FALSE} \newline
  \indented{return TRUE} \newline
    end if \newline
    \comment{The following code considers the remaining case where $Df(x_0) \neq 0$ and $Df(x_1) \neq 0$.} \newline
  $\displaystyle A := Df(x_0){x_1 - x_0 \over f(x_1) - f(x_0)}$ \newline
  $\displaystyle B := Df(x_1) {x_1 - x_0 \over f(x_1) - f(x_0)}$ \newline
    \comment{The variables $A$ and $B$ correspond directly to the theoretical foundation for positive quartic polynomials established in \cite{ulrich1994positivity}, first defined after Equation (18).} \newline
  $\displaystyle \gamma_0 := 4 {Df(x_0) \over Df(x_1)} (B/A)^{3/4}$ \newline
  $\displaystyle \gamma_1 := {x_1 - x_0 \over Df(x_1)} (B/A)^{3/4}$ \newline
  $\alpha_0 := 4 (B/A)^{1/4}$ \newline
  $\displaystyle \alpha_1 := -{x_1 - x_0 \over Df(x_1)} (B/A)^{1/4}$ \newline 
  $\displaystyle \beta_0 := 30 - {12 \big(Df(x_0) + Df(x_1)\big) (x_1 - x_0) \over \big(f(x_1) - f(x_0)\big) \sqrt{A}\sqrt{B}}$ \newline
  $\displaystyle \beta_1 := {-3 (x_1 - x_0)^2 \over 2 \big(f(x_1) - f(x_0)\big) \sqrt{A} \sqrt{B}} $ \newline
    \comment{The $\gamma$, $\alpha$, and $\beta$ terms with subscripts $0$ and $1$ are algebraic reductions of the simplified conditions for satisfying Theorem 2 in Equation (16) of \cite{ulrich1994positivity}. These terms with subscripts $0$ and $1$ make the computation of $\alpha$, $\beta$, and $\gamma$ functions of the second derivative terms, as seen in Step \ref{alg1-abg} below.} \newline
  $\gamma := \gamma_0 + \gamma_1 D^2f(x_0)$ \newline
  $\alpha := \alpha_0 + \alpha_1 D^2f(x_1)$ \newline
  $\beta := \beta_0 + \beta_1 \big(D^2f(x_0) - D^2f(x_1)\big)$ \newline
  if $(\beta \leq 6)$ return $\big(\alpha > - (\beta + 2) / 2\big)$ \newline
  else $\qquad$ return $\big(\gamma > -2 \sqrt{\beta - 2}\big)$ \newline
  end if \newline
}
\endnumberedlist
\vskip 2mm
\hrule
\vskip 3mm
\parindent=10pt

%% ----------------------------------------------------------------------

The reason for structuring the $\alpha$, $\beta$, and $\gamma$ computations in terms of the second derivative of $f$ (seen in Step \ref{alg1-abg} of Algorithm \ref{alg:check-monotone}) will become more apparent later. The next problem to consider is that of making a nonmonotone order six polynomial piece into a monotone one by modifying its first and second derivative values at the ends of an interval. Note that the actual value of the function at the ends of the interval is not modified, as the resulting polynomial needs to interpolate.

Notice that both Algorithms \ref{alg:check-monotone} and \ref{alg:make-monotone} have ${\cal O}(1)$ runtime assuming a fixed level of precision is chosen beforehand. A constant number of operations are needed for verifying monotonicity (Algorithm \ref{alg:check-monotone}), while a constant set of operations and a single binary search are performed for enforcing monotonicity (Algorithm \ref{alg:make-monotone}). The search requires a fixed number of steps to achieve any predetermined relative precision, since its accuracy is predetermined and not a function of the problem at hand. Also note that an efficient implementation of Algorithm \ref{alg:make-monotone} only needs to recalculate Steps \ref{alg1-abg} and \ref{alg1-last} of Algorithm \ref{alg:check-monotone} during the binary search. Next, an algorithm for constructing a monotone quintic spline interpolant is presented.

%% ----------------------------------------------------------------------

\vskip 5mm
\thickrule
{\bf Algorithm 2:} {\tt monotone\_spline}$\big((k_1$, $\ldots$, $k_n)$, $f$, $s\big)$
where $(k_1$, $\ldots$, $k_n)$ is an increasing sequence of real
numbers, $f$ is an order six piecewise polynomial with breakpoints
$k_1$, $\ldots$, $k_n$ defined by the data
$\bigl\{f(k_i)\bigr\}_{i=1}^n$, $\bigl\{Df(k_i)\bigr\}_{i=1}^n$,
$\bigl\{D^2f(k_i)\bigr\}_{i=1}^n$, and $s > 1$ is an integer shrink
factor.
\hrule
\vskip 3mm
    \newline create queue
      \newline create changed$(1$, $\ldots$, $n) := $ FALSE
      \newline create step\_size$(1$, $\ldots$, $n) := \bigl(Df(k_1)/s$, $\ldots$, $Df(k_n)/s\bigr)$
      \comment{The three variables defined in Step \ref{step:create} are used to ensure the eventual achievement of monotonicity. The {\tt queue} is a standard first in first out (FIFO) queue with {\tt enqueue} and {\tt dequeue} operations. The array {\tt changed} contains Booleans describing whether or not a breakpoint belongs to an interval that has been modified to enforce monotonicity. The {\tt step\_size} array contains the real-valued derivative decrement step sizes to use in the search for a valid monotone spline.}
    \newline for $i := 1$, $\ldots$, $n-1$
      \indented{enqueue $\bigl((k_i, k_{i+1})\bigr)$}
      end for
      \comment{Initially, all intervals must be checked for monotonicity. The following loop will continue until all intervals are verified as monotone without need for modification.}
    \newline while $($size$($queue$) > 0)$
    \newline \indented{$(k_j$, $k_{j+1}) := $ dequeue}
    \newline \indented{if $\big($not is\_monotone$(k_j$, $k_{j+1}$, $f)\big)$}
    \newline \indented{\indented{if $\big($changed$(j)\big) \qquad \ Df(k_j) := Df(k_j) -{}$step\_size$(j)$}}
      \indented{\indented{if $\big($changed$(j+1)\big) \quad Df(k_{j+1}) := Df(k_{j+1}) -{}$step\_size$(j+1)$}}
      \indented{\indented{\comment{If a breakpoint belongs to an interval that has been previously modified, then the enforced monotonicity conditions on the second derivatives of adjacent intervals must contradict one another. In turn, the involved first derivatives are decreased towards zero by a predetermined step size.}}}
    \newline \indented{\indented{make\_monotone$(k_j,\ k_{j+1},\ f)$}}
    \newline \indented{\indented{changed$(j) :=$ TRUE}}
      \indented{\indented{changed$(j+1) :=$ TRUE}}
    \newline \indented{\indented{if $\big(j > 1$ and $(k_{j-1}, k_j) \not\in$ queue$\big)$ enqueue $\bigl((k_{j-1}, k_j)\bigr)$}}
      \indented{\indented{enqueue $\bigl((k_j, k_{j+1})\bigr)$}}
      \indented{\indented{if $\big(j+1 < n$ and $(k_{j+1}, k_{j+2}) \not\in$ queue$\big)$ enqueue $\bigl((k_{j+1}, k_{j+2})\bigr)$}}
      \indented{\indented{\comment{Step \ref{step:changed} records the endpoints of the current interval as having been changed, while \ref{step:requeue} adds adjacent intervals to {\tt queue} that may have inadvertently been made nonmonotone by the changes to the present interval.}}}      
      \indented{end if}
      end while
\vskip 2mm
\hrule
\vskip 3mm

%% ----------------------------------------------------------------------

It is mentioned in \cite{ulrich1994positivity} that for sufficiently small $Df(k_i)$ and $Df(k_{i+1})$ the admissible solution interval of second derivative values becomes arbitrarily large. It can also be seen that decreasing the assigned derivative values towards zero will eventually cause Steps \ref{alg1-alpha} through \ref{alg1-gamma} of Algorithm \ref{alg:check-monotone} to all fail, resulting in Step \ref{alg1-simplified-true} returning {\tt TRUE}.

If successive monotonicity fixes are required for all intervals, the worst case runtime of Algorithm \ref{alg:monotone-spline} is ${\cal O}(s n)$ for $n$ breakpoints and shrink factor $s$. In practice this worst case has been observed to be unlikely.

\heading{3. EXPERIMENTAL RESULTS}

Given the increased order of approximation, it is naturally expected that some accuracy benefit could be gained by using a quintic spline over a cubic spline. Experiments 1 and 2 below test some accuracy differences between monotone cubic and quintic splines. Experiment 3 analyzes the number of operations performed by Algorithm \ref{alg:monotone-spline} with increasingly large samples of monotone data.

\heading{3.1 \enspace Approximating a Trigonometric Function}

For this experiment, the function $\sin(x) + x$ is considered over the interval $[0,(5/2)\pi]$ as seen in Figure \ref{fig:cubic_quintic_sin}. Given an increasing number of points, the maximum error of cubic and quintic approximations is shown in Figure \ref{fig:experiment_1}. The quintic approximation consistently has a maximum error that is roughly half that of the cubic approximation, given the same number of points. There is an increase in the error gap between the two approximations as more points are added. The quintic spline requires only $700$ points to approximate the function to the same accuracy achieved by the cubic with $1000$ points.

%% \begin{figure}[h]
%%   \centering
%%   \includegraphics[width=.7\textwidth]{cubic-quintic-sin}
%%   \caption{Depicted above are the monotone cubic and quintic spline interpolants of the function $\sin(x) + x$ over the points $[0, (5/2) \pi]$.  Notice that the maximum error of the cubic interpolant is larger, because it only captures first derivative information at the points. The quintic interpolant captures both first and second derivative information at the points.}
%% \end{figure}

%% \begin{figure}
%%   \centering
%%   \includegraphics[width=.7\textwidth]{experiment_1_errors}
%%   \caption{The error in the quintic and cubic approximations of $\sin(x) + x$ with increasing number of points. Both axes are log scaled, where $\mu$ means $\times 10^{-6}$ and n means $\times 10^{-9}$ on the vertical axis. The quintic approximation generally provides an approximation with about half of the maximum error of the cubic approximation, while the relative gap between grows with increasing number of points. In order to achieve the same absolute error the quintic spline requires ten fewer points at max error $10^{-4}$, thirty fewer points at max error $10^{-6}$, and $300$ fewer points at max error $10^{-8}$.}
%% \end{figure}

\heading{3.2 \enspace Random Number Generation}

Consider the cumulative distribution function (CDF) defined by a mixture of three Gaussian distributions and visualized in Figure \ref{fig:experiment_2_distribution}. The distributions have weights $(.3, .6, .1)$, means $(.2, .45, .85)$ and standard deviations $(.05, .08, .03)$. $200$ random samples are generated from this distribution with both cubic and quintic approximations from four points, and over $100$ trials the resulting spread of empirical distribution functions (EDFs) are visualized in Figure \ref{fig:experiment_2_results}.

While an increased number of points could be used to decrease the error in either approximation, the purpose of this experiment is to demonstrate the effect approximation error has on random number generation. The errors in the CDFs are magnified by the variance involved in randomly sampling from a distribution. As a result, while the cubic and quintic spline approximations have similar maximum error, the cubic distribution is significantly distorted around the first and second modes.

>>>>

% Algorithm XXXX: MQSI---Monotone quintic spline interpolation,
% Thomas C.H. Lux, Layne T. Watson, Tyler H. Chang,
% and William Thacker
% ACM Trans. Math. Software, ?/?/2020.
% 
%
\magnification=1000 \hsize=30pc \vsize=47pc
\parindent=10pt \parskip=0pt plus 3pt
\baselineskip=13pt plus 2pt minus 1pt \lineskiplimit=2pt
\lineskip=2pt plus 2pt %\raggedright
\tolerance=800
%\abovedisplayskip=6pt plus 6pt minus 3pt
%\abovedisplayshortskip=0pt plus 3pt
%\belowdisplayskip=6pt plus 6pt minus 3pt
%\belowdisplayshortskip=6pt plus 2pt minus 3pt

% \cite is a placeholder to use for references while writing.
\def\cite#1{[{\tt REF}]}

@@

% Commands for inserting an 'algorithm' section.
\def\thickrule{\hrule\hrule\hrule}
\def\codent{\hskip 1.8em} % <- 'tab' for indentation.

@@

%Postscript figures.
\input epsf
\def\lcaption#1#2{{\noindent\rmVIII Fig.~#1. #2\par}}
\def\ltabcaption#1#2{{\noindent\rmVIII Table #1. #2\par}}

@@

\def\leftrh{Lux et al.}
\def\rightrh{Algorithm XXXX: MQSI}

{\baselineskip=24pt
\leftline{\hlvXVIII Algorithm XXXX: MQSI---Monotone Quintic}
\leftline{\hlvXVIII Spline Interpolation}}
\bigskip\medskip
\leftline{\hlv THOMAS C. H. LUX and LAYNE T. WATSON}
\smallskip
\leftline{\hlv Virginia Polytechnic Institute and State University}
\smallskip
\leftline{\hlv TYLER H. CHANG}
\smallskip
\leftline{\hlv Argonne National Laboratory}
\smallskip
\leftline{\hlv WILLIAM I. THACKER}
\smallskip
\leftline{\hlv Winthrop University}
\bigskip

\hrule\bigskip\smallskip \footnote{}{\hskip
-\parindent{\parindent=0pt\rmVIII Authors' addresses:
T. C. H. Lux, L. T. Watson, Departments of Computer Science, Mathematics,
and Aerospace and Ocean Engineering, Virginia Polytechnic Institute \&
State University, Blacksburg, VA 24061; e-mails: {\ttVIII tchlux@vt.edu,
ltwatson@computer.org}; T. H. Chang, Mathematics and Computer Science
Division, Argonne National Laboratory, 9700 South Cass Avenue, Bldg. 240,
Lemont, IL 60439; e-mail: {\ttVIII thchang@vt.edu}; W. I. Thacker, Winthrop
University, Rock Hill, SC 47405; {\ttVIII thackerw@winthrop.edu}.

\xpar
% Permission to make digital/hard copy of
%part or all of this work for personal or classroom use is granted
%without fee provided that the copies are not made or distributed for
%profit or commercial advantage, the copyright notice, the title of
%the publication, and its date appear, and notice is given that
%copying is by permission of the ACM, Inc. To copy otherwise, to
%republish, to post on servers, or to redistribute to lists, requires
%specific permission and/or fee. \xpar\copyright\ 2018 by the
%Association for Computing Machinery, Inc. \xpar 
}}

% Abstract and Introduction
{\rmVIII\parindent=0pt MQSI is a Fortran 2003 subroutine for constructing
monotone quintic spline interpolants to monotone data. Using sharp
theoretical monotonicity constraints, first and second derivative estimates
at data provided by a quadratic facet model are refined to produce a $C^2$
monotone interpolant.  Algorithm and implementation details, complexity
and sensitivity analyses, usage information, and a brief performance study
are included.

@@

%% Section 1
\input 1-introduction

%% Section 2
\input 2-monotone-quintic

%% Section 3
\input 3-spline-rep

%% Section 4
\input 4-analysis

%% Section 5
\input 5-applications

%% Acknowledgements
%% \heading{Acknowledgments}
%% This work was supported by the National Science Foundation Grants CNS-1565314 and CNS-1838271.

% ====================================================================
% Bibliography
\heading{\hlvVIII BIBLIOGRAPHY} {
\baselineskip=2pt \parskip=100mm
\frenchspacing
% Temporary 'caps', 'sl', and 'rm'.
\let\caps=\capsVIII
\let\sl=\slVIII
\let\rm=\rmeight \rmVIII

% \ys makes a half-em horizontal space.
\def\ys{\hskip 1em minus .5em}

%\refj is for regular journal articles
%  #1 -- Author(s)
%  #2 -- Year
%  #3 -- Title
%  #4 -- Journal
%  #5 -- Volume, number, and pages
\def\refj#1#2#3#4#5{\noindent\hangindent=10pt\hangafter=1
{\caps #1}\ys #2.\ys {\rm #3}. {\sl #4}, #5.\vskip 6pt}

%\refb is for books
%  #1 -- Author(s)
%  #2 -- Year
%  #3 -- Title
%  #4 -- Publisher, City, State
\def\refb#1#2#3#4{\noindent\hangindent=10pt\hangafter=1
{\caps #1}\ys #2.\ys{\sl #3}. {\rm #4}.\vskip 6pt}

%\reft is for tech. reports, theses, and conferences
%  #1 -- Author(s)
%  #2 -- Year
%  #3 -- Title
%  #4 -- Technical report ID, Publisher, City, State
\def\reft#1#2#3#4{\noindent\hangindent=10pt\hangafter=1
{\caps #1}\ys #2.\ys{\rm #3}. {\rm #4}.\vskip 6pt}

%% -------------------------------------------------------------------

% berglund2009planning
\refj{Berglund, T., Brodnik, A., Jonsson, H., Staffanson, M., and
  Soderkvist, I.}{2009}{Planning smooth and obstacle-avoiding B-spline
  paths for autonomous mining vehicles}{IEEE Transactions on
  Automation Science and Engineering}{7(1):167--172}

% brennan2019measure
\refj{Brennan, A.}{2020}{Measure, Modulation and Metadesign: NC
  Fabrication in Industrial Design and Architecture}{Journal of Design
  History}{33(1):66--82}

% fritsch1980monotone
\refj{Fritsch., F and Carlson., R.}{1980}{Monotone piecewise cubic
  interpolation}{SIAM Journal on Numerical Analysis}{17(2):238–-246}

% fritsch1982LLNL
\reft{Fritsch, F. N.}{1982}{Piecewise cubic Hermite interpolation
  package, final specifications}{Computer Documentation UCID-30194,
  Lawrence Livermore National Laboratory}

% gregory1985shape
\reft{Gregory, J. A.}{1985}{Shape preserving spline
  interpolation}{19850020252, Brunel University}

% haralick1981facet
\refj{Haralick, R. M., and Watson, L. T.}{1981}{A facet model for
  image data.}{Computer Graphics and Image Processing}{15(2):113--129}

% he1998monotone
\refj{He, X. and Shi, P.}{1998}{Monotone B-spline smoothing}{Journal
  of the American Statistical Association}{93(442):643--650}

% herman2015techniques
\reft{Herman, D. L., and Oftedal, M. J.}{2015}{Techniques and
  workflows for computer graphics animation systems}{US Patent
  9,216,351}

% hess1994positive
\refj{He{\ss}, W. and Schmidt, J. W.}{1994}{Positive quartic, monotone
  quintic $C^2$-spline interpolation in one and two dimensions}{Journal
  of Computational and Applied Mathematics}{55(1):51–-67}

% knott2012interpolating
\reft{Knott, G. D.}{2012}{Interpolating cubic splines}{Springer Science
  \& Business Media, Volume 18}

% leitenstorfer2006generalized
\refj{Leitenstorfer, F. and Tutz, G.}{2006}{Generalized monotonic
  regression based on B-splines with an application to air pollution
  data}{Biostatistics}{8(3):654–-673}

% lux2020thesis
\reft{Lux, T. C. H.}{2020}{Interpolants and error bounds for modeling
  and predicting variability in computer systems}{Thesis, Virginia
  Polytechnic Institute and State University, Blacksburg, Virginia}

% piah2011improved
\refj{Piah, A. R. M. and Unsworth, K.}{2011}{Improved sufficient
  conditions for monotonic piecewise rational quartic
  interpolation}{Sains Malaysiana}{40(10):1173–-1178}

% quint2003scalable
\refj{Quint, A.}{2003}{Scalable vector graphics}{IEEE
  MultiMedia}{10(3):99–102}

% ramsay1988monotone
\refj{Ramsay, J. O.}{1988}{Monotone regression splines in
  action}{Statistical Science}{3(4):425-–441}

% ulrich1994positivity
\refj{Ulrich, G. and Watson, L. T.}{1994}{Positivity conditions for
  quartic polynomials}{SIAM Journal on Scientific
  Computing}{15(3):528–-544}

% wang2004rational
\refj{Wang, Q. and Tan, J.}{2004}{Rational quartic spline involving
  shape parameters}{Journal of Information and Computational
  Science}{1(1):127–-130}

% yao2018unconditionally
\refj{Yao, J. and Nelson, K. E.}{2018}{An Unconditionally Monotone
  $C^2$ Quartic Spline Method with Nonoscillation
  Derivatives}{Advances in Pure Mathematics}{8(LLNL-JRNL-742107)}

%% -------------------------------------------------------------------
